---
title: "Hierarchical Variable Clustering with HClustVar"
author: "Eugénie Barlet, Modou Mboup, Axel Cano"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchical Variable Clustering with HClustVar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

### What is Variable Clustering?

Unlike traditional clustering that groups **observations** (rows), variable clustering groups **variables** (columns) based on their similarity. This technique is particularly useful for:

- **Feature selection**: Identify redundant variables within clusters
- **Data understanding**: Discover the underlying structure of your dataset
- **Multicollinearity reduction**: Group highly correlated predictors in regression models

### The algortihm

#### HClustVar

The `HClustVar` class implements hierarchical clustering of variables using:

- **Quantitative variables**: Based on correlation (Pearson's r or r²)
- **Qualitative variables**: Based on Cramér's V (association measure)
- **Mixed variables**: Automatic discretization and Cramér's V
- **Multiple linkage methods**: Ward, single, complete, average, etc.

### Key Features

✓ Automatic variable type detection  
✓ Flexible distance metrics  
✓ Cluster quality diagnostics (dendrogram, cohesion, silhouette)  
✓ Prediction for new variables  
✓ Computation of cluster centroids (synthetic variables)  

---

## Installation

```{r eval = FALSE}
# Install from GitHub (or your package repository)
devtools::install_github("username/sise.var.clustering")
```

```{r}
# Load the package
library(sise.var.clustering)
```

---

## Quick Start: Complete Workflow

Here's a minimal example to get you started:

```{r eval = FALSE}

df <- read.csv("data/players_22_cleaned.csv")

# 1. Create the model
hc <- HClustVar$new(vartype = "quant", dist.metric = "rsquare")

# 2. Fit on your data
hc$fit(my_data)

# 3. Find optimal number of clusters
hc$plot_cohesion()

# 4. Cut the tree
hc$cut_tree(k = 3)

# 5. Visualize results
hc$plot_silhouette()

# 6. Get detailed statistics
results <- hc$summary()
print(results$clust_summary)
```

---

## Example 1: Quantitative Variables

### Dataset Creation

Let's create a dataset with 3 groups of correlated variables:

```{r}
set.seed(123)
n <- 100

# Group 1: Economic indicators (correlated)
gdp <- rnorm(n, mean = 50000, sd = 10000)
income <- gdp * 0.7 + rnorm(n, sd = 2000)
consumption <- income * 0.6 + rnorm(n, sd = 1000)

# Group 2: Health indicators (correlated)
life_expectancy <- rnorm(n, mean = 75, sd = 5)
infant_mortality <- 100 - life_expectancy + rnorm(n, sd = 2)
healthcare_spending <- life_expectancy * 50 + rnorm(n, sd = 200)

# Group 3: Education indicators (correlated)
literacy_rate <- rnorm(n, mean = 85, sd = 10)
school_enrollment <- literacy_rate * 0.9 + rnorm(n, sd = 5)
university_rate <- literacy_rate * 0.5 + rnorm(n, sd = 8)

# Combine into dataframe
data_quant <- data.frame(
  gdp, income, consumption,
  life_expectancy, infant_mortality, healthcare_spending,
  literacy_rate, school_enrollment, university_rate
)

head(data_quant)
```

### Model Creation and Fitting

```{r}
# Initialize the model
hc_quant <- HClustVar$new(
  vartype = "quant",
  dist.metric = "rsquare",  # Use squared correlation
  cah.method = "ward.D"      # Ward's method for homogeneous clusters
)

# Fit the model
hc_quant$fit(data_quant)

# Display model information
print(hc_quant)
```

### Selecting the Number of Clusters

#### Method 1: Dendrogram Visualization

```{r}
hc_quant$plot_dendrogram()
```

**Interpretation**: The dendrogram shows the hierarchical structure. Variables that merge at lower heights are more similar.

#### Method 2: Cohesion Gain Curve (Elbow Method)

```{r}
hc_quant$plot_cohesion()
```

**Interpretation**: Look for the "elbow" where the curve flattens. Here, **k = 3** appears optimal as the gain diminishes significantly after this point.

### Cutting the Tree

Based on the diagnostics, we choose **k = 3 clusters**:

```{r}
hc_quant$cut_tree(k = 3)
```

### Assessing Cluster Quality

#### Silhouette Plot

```{r}
hc_quant$plot_silhouette()
```

**Interpretation**:
- **Positive values** (most bars): Variables are well assigned
- **Negative values**: Potential misclassification
- **Average silhouette width > 0.5**: Good clustering quality

#### Summary Statistics

```{r}
results <- hc_quant$summary()

# Cluster-level summary
print(results$clust_summary)
```

**Key Metrics**:
- `n_members`: Number of variables in each cluster
- `var_explained`: Variance explained by the cluster centroid
- `prop_explained`: Average variance explained per variable (higher = more homogeneous)

```{r}
# Variable-level details
print(results$clust_members)
```

**Key Metrics**:
- `own_cluster_R2`: How well the variable fits its cluster (> 0.7 is good)
- `next_closest_R2`: Similarity with the nearest other cluster
- `1 - R2_ratio`: Quality indicator (< 0.3 is well classified)

### Interpreting Clusters

```{r}
# Which variables are in each cluster?
for (i in 1:3) {
  cat(sprintf("\nCluster %d:\n", i))
  cluster_vars <- names(hc_quant$labels)[hc_quant$labels == i]
  cat(paste("  -", cluster_vars, collapse = "\n"))
}
```

**Expected grouping**:
- Cluster 1: Economic variables (GDP, income, consumption)
- Cluster 2: Health variables (life expectancy, mortality, healthcare)
- Cluster 3: Education variables (literacy, enrollment, university)

### Using Cluster Centroids

The centroids are synthetic variables (first principal components) that summarize each cluster:

```{r}
# Access centroids
centroids <- hc_quant$centroids
head(centroids)

# Use centroids instead of original variables (dimensionality reduction)
# For example, in a regression model:
# model <- lm(y ~ centroids[, 1] + centroids[, 2] + centroids[, 3])
```

---

## Example 2: Qualitative Variables

### Dataset Creation

```{r}
set.seed(456)
n <- 150

# Group 1: Demographics
gender <- factor(sample(c("Male", "Female"), n, replace = TRUE))
age_group <- factor(sample(c("18-30", "31-50", "51+"), n, replace = TRUE))

# Group 2: Location (correlated with demographics)
region <- factor(sample(c("North", "South", "East", "West"), n, replace = TRUE))
urban_rural <- factor(sample(c("Urban", "Rural"), n, replace = TRUE, 
                             prob = c(0.7, 0.3)))

# Group 3: Preferences (associated with demographics)
product_preference <- factor(sample(c("A", "B", "C"), n, replace = TRUE))
brand_loyalty <- factor(sample(c("Low", "Medium", "High"), n, replace = TRUE))

data_qual <- data.frame(
  gender, age_group, region, urban_rural, 
  product_preference, brand_loyalty
)

head(data_qual)
```

### Model Fitting

```{r}
# Initialize for qualitative variables
hc_qual <- HClustVar$new(
  vartype = "qual",     # Qualitative variables
  cah.method = "ward.D"
)

# Fit the model
hc_qual$fit(data_qual)

# Diagnostic plots
hc_qual$plot_dendrogram()
```

```{r}
hc_qual$plot_cohesion()
```

### Cutting and Analysis

```{r}
# Cut into 2 clusters
hc_qual$cut_tree(k = 2)

# Silhouette analysis
hc_qual$plot_silhouette()

# Summary
results_qual <- hc_qual$summary()
print(results_qual$clust_summary)
print(results_qual$clust_members)
```

**Note**: For qualitative variables, the similarity metric is **Cramér's V** (ranges from 0 to 1), which measures association between categorical variables.

---

## Example 3: Mixed Variables (Auto-Detection)

### Dataset Creation

```{r}
set.seed(789)
n <- 120

# Quantitative variables
age <- rnorm(n, mean = 40, sd = 12)
income <- rnorm(n, mean = 50000, sd = 15000)
satisfaction_score <- rnorm(n, mean = 7, sd = 2)

# Qualitative variables
education <- factor(sample(c("High School", "Bachelor", "Master", "PhD"), 
                           n, replace = TRUE))
employment <- factor(sample(c("Full-time", "Part-time", "Unemployed"), 
                            n, replace = TRUE))
marital_status <- factor(sample(c("Single", "Married", "Divorced"), 
                                n, replace = TRUE))

data_mixed <- data.frame(
  age, income, satisfaction_score,
  education, employment, marital_status
)

head(data_mixed)
```

### Auto-Detection and Fitting

```{r}
# Use 'auto' to automatically detect variable types
hc_auto <- HClustVar$new(
  vartype = "auto",      # Automatic detection
  cah.method = "ward.D"
)

# Fit the model (detection happens here)
hc_auto$fit(data_mixed)

# Check detected type
cat("Detected variable type:", hc_auto$vartype, "\n")
```

**Note**: With mixed data, quantitative variables are automatically discretized into quantiles, and all variables are treated as qualitative using Cramér's V.

### Analysis

```{r}
# Diagnostics
hc_auto$plot_cohesion()
```

```{r}
# Cut tree
hc_auto$cut_tree(k = 2)

# Silhouette
hc_auto$plot_silhouette()

# Summary
results_mixed <- hc_auto$summary()
print(results_mixed$clust_summary)
```

---

## Example 4: Predicting New Variables

### Scenario

You've built a clustering model and want to assign **new variables** to existing clusters without refitting the entire model.

```{r}
# Use the quantitative model from Example 1
# (already fitted and cut)

# Create new variables
new_variables <- data.frame(
  poverty_rate = 100 - data_quant$income / 1000 + rnorm(n, sd = 5),
  doctor_density = data_quant$healthcare_spending / 100 + rnorm(n, sd = 10),
  research_investment = data_quant$university_rate * 1000 + rnorm(n, sd = 500)
)

head(new_variables)
```

### Prediction

```{r}
# Predict cluster membership
predictions <- hc_quant$predict(new_variables)
print(predictions)
```

**Interpretation**: 
- `poverty_rate` → Likely assigned to economic cluster (Cluster 1)
- `doctor_density` → Likely assigned to health cluster (Cluster 2)
- `research_investment` → Likely assigned to education cluster (Cluster 3)

### Validation

```{r}
# Check correlations with cluster centroids
for (i in 1:ncol(new_variables)) {
  var_name <- colnames(new_variables)[i]
  assigned_cluster <- predictions[i]
  
  # Correlation with assigned cluster centroid
  cor_assigned <- cor(new_variables[[i]], hc_quant$centroids[, assigned_cluster])
  
  cat(sprintf("%s → Cluster %d (R = %.2f)\n", 
              var_name, assigned_cluster, cor_assigned))
}
```

---

## Advanced Topics

### Choosing the Distance Metric

For **quantitative variables**, two metrics are available:

#### 1. Correlation (`dist.metric = "r"`)

```{r}
hc_r <- HClustVar$new(vartype = "quant", dist.metric = "r")
```

**Use when**: You want to distinguish between positively and negatively correlated variables.

- Distance = $\sqrt{1 - r}$
- Variables with $r = -1$ (perfect negative correlation) are **far apart**

#### 2. Squared Correlation (`dist.metric = "rsquare"`)

```{r}
hc_r2 <- HClustVar$new(vartype = "quant", dist.metric = "rsquare")
```

**Use when**: You care about the strength of association, regardless of direction.

- Distance = $\sqrt{1 - r^2}$
- Variables with $r = -1$ or $r = +1$ are **equally close**

**Example comparison**:

```{r}
# Create two variables: one positively, one negatively correlated with X
X <- rnorm(100)
Y_pos <- X + rnorm(100, sd = 0.5)      # Positive correlation
Y_neg <- -X + rnorm(100, sd = 0.5)     # Negative correlation

test_data <- data.frame(X, Y_pos, Y_neg)

# With 'r' metric
hc_r <- HClustVar$new(vartype = "quant", dist.metric = "r")
hc_r$fit(test_data)
hc_r$plot_dendrogram()
title(main = "Metric = 'r' (separates positive/negative)")

# With 'rsquare' metric
hc_r2 <- HClustVar$new(vartype = "quant", dist.metric = "rsquare")
hc_r2$fit(test_data)
hc_r2$plot_dendrogram()
title(main = "Metric = 'rsquare' (groups by strength)")
```

### Choosing the Linkage Method

Different `cah.method` options produce different clustering behaviors:

```{r eval = FALSE}
# Ward's method (default) - minimizes within-cluster variance
hc_ward <- HClustVar$new(cah.method = "ward.D")

# Single linkage - clusters based on closest members (can create chains)
hc_single <- HClustVar$new(cah.method = "single")

# Complete linkage - clusters based on farthest members (compact clusters)
hc_complete <- HClustVar$new(cah.method = "complete")

# Average linkage - balanced approach
hc_average <- HClustVar$new(cah.method = "average")
```

**Recommendation**: Start with `"ward.D"` (default) as it generally produces the most homogeneous clusters.

### Dealing with Large Datasets

For datasets with many variables (> 50), `plot_cohesion()` can be slow as it tests all possible numbers of clusters:

```{r eval = FALSE}
# Strategy 1: Inspect dendrogram first to estimate k
hc$plot_dendrogram()

# Strategy 2: Cut at a specific height instead of testing all k values
hc$cut_tree(h = 1.5)  # Cut at height 1.5

# Strategy 3: Use a subset for initial exploration
hc_subset <- HClustVar$new(vartype = "quant")
hc_subset$fit(data[, 1:20])  # Test on first 20 variables
hc_subset$plot_cohesion()
```

---

## Practical Tips and Best Practices

### 1. Data Preparation

```{r eval = FALSE}
# For quantitative variables: check for outliers
# Extreme outliers can distort correlations
summary(data_quant)
boxplot(data_quant)

# For qualitative variables: ensure all are factors
data_qual[] <- lapply(data_qual, as.factor)

# For mixed data: be explicit about types
data_mixed$age <- as.numeric(data_mixed$age)
data_mixed$education <- as.factor(data_mixed$education)
```

### 2. Interpreting Results

**Good cluster quality indicators**:
- Silhouette width > 0.5
- `own_cluster_R2` > 0.7 for most variables
- `prop_explained` > 0.7 for most clusters
- Clear "elbow" in cohesion plot

**Warning signs**:
- Negative silhouette values
- `1 - R2_ratio` > 0.7 (potential misclassification)
- Very unbalanced cluster sizes
- No clear elbow in cohesion plot

### 3. Dimensionality Reduction Workflow

```{r eval = FALSE}
# 1. Cluster variables
hc <- HClustVar$new(vartype = "quant", dist.metric = "rsquare")
hc$fit(original_data)
hc$cut_tree(k = 5)

# 2. Extract centroids (5 synthetic variables instead of, say, 50 originals)
reduced_data <- as.data.frame(hc$centroids)
colnames(reduced_data) <- paste0("Cluster_", 1:5)

# 3. Use in downstream analysis
model <- lm(target ~ ., data = cbind(target = y, reduced_data))
```

### 4. Variable Selection Within Clusters

```{r eval = FALSE}
# Select the most representative variable from each cluster
results <- hc$summary()

for (k in 1:hc$n_clusters) {
  cluster_vars <- results$clust_members[results$clust_members$cluster == k, ]
  best_var <- rownames(cluster_vars)[which.max(cluster_vars$own_cluster_R2)]
  
  cat(sprintf("Cluster %d best representative: %s (R² = %.2f)\n",
              k, best_var, max(cluster_vars$own_cluster_R2)))
}
```

---

## Comparison with Other Methods

### HClustVar vs. K-means (for observations)

| Feature | HClustVar | K-means |
|---------|-----------|---------|
| **Clusters** | Variables (columns) | Observations (rows) |
| **Goal** | Group related features | Group similar individuals |
| **Output** | Variable clusters + centroids | Observation clusters + centers |
| **Use case** | Reduce dimensionality, understand variables | Segment population, find groups |

### HClustVar vs. PCA

| Feature | HClustVar | PCA |
|---------|-----------|-----|
| **Interpretability** | High (clusters of original variables) | Medium (abstract components) |
| **Variable selection** | Possible (choose one per cluster) | Automatic (all contribute to PCs) |
| **Qualitative data** | Supported | Requires dummy coding |
| **Flexibility** | Multiple distance metrics | Fixed (variance-based) |

**When to use HClustVar**: When you need interpretable groups of variables and want to select/reduce features while maintaining their original meaning.

**When to use PCA**: When you need maximum variance explained and don't mind abstract components.

---

## Troubleshooting

### Error: "Model must be fitted before cutting"

```{r eval = FALSE}
hc <- HClustVar$new()
hc$cut_tree(k = 3)  # ❌ Error!

# Solution: Fit first
hc$fit(data)
hc$cut_tree(k = 3)  # ✓ OK
```

### Error: "new_data must have same number of rows"

```{r eval = FALSE}
hc$fit(data)  # 100 rows
new_data <- data.frame(x = rnorm(50))  # 50 rows
hc$predict(new_data)  # ❌ Error!

# Solution: Ensure same number of observations
new_data <- data.frame(x = rnorm(100))  # 100 rows
hc$predict(new_data)  # ✓ OK
```

### Warning: "Contingency table too small for chi-square test"

This occurs with qualitative variables that have very few categories or observations.

```{r eval = FALSE}
# Problematic data
tiny_data <- data.frame(
  var1 = factor(c("A", "A", "B")),  # Only 3 observations
  var2 = factor(c("X", "Y", "X"))
)

# Solution: Use more observations or combine rare categories
```

### No clear elbow in cohesion plot

```{r eval = FALSE}
# Try different linkage methods
hc_ward <- HClustVar$new(cah.method = "ward.D")
hc_complete <- HClustVar$new(cah.method = "complete")
hc_average <- HClustVar$new(cah.method = "average")

# Compare dendrograms
hc_ward$fit(data)
hc_ward$plot_dendrogram()

hc_complete$fit(data)
hc_complete$plot_dendrogram()
```

---

## Real-World Example: Marketing Survey Data

Let's analyze a realistic marketing dataset with customer survey responses:

```{r}
set.seed(2024)
n <- 200

# Customer demographics
age <- rnorm(n, 35, 10)
income <- rnorm(n, 60000, 20000)

# Product satisfaction scores (1-10)
product_quality <- sample(1:10, n, replace = TRUE)
price_satisfaction <- 11 - product_quality + sample(-2:2, n, replace = TRUE)
delivery_speed <- sample(1:10, n, replace = TRUE)

# Service ratings (1-10)
customer_service <- sample(1:10, n, replace = TRUE)
technical_support <- customer_service + sample(-2:2, n, replace = TRUE)
response_time <- customer_service + sample(-2:2, n, replace = TRUE)

# Loyalty indicators (1-10)
recommendation_score <- (product_quality + customer_service) / 2 + 
  sample(-2:2, n, replace = TRUE)
repurchase_intent <- recommendation_score + sample(-2:2, n, replace = TRUE)
brand_trust <- (product_quality + customer_service + delivery_speed) / 3 + 
  sample(-2:2, n, replace = TRUE)

# Combine
survey_data <- data.frame(
  age, income,
  product_quality, price_satisfaction, delivery_speed,
  customer_service, technical_support, response_time,
  recommendation_score, repurchase_intent, brand_trust
)

# Clip values to 1-10 range for ratings
survey_data[, 3:11] <- apply(survey_data[, 3:11], 2, function(x) {
  pmax(1, pmin(10, x))
})

head(survey_data)
```

### Analysis

```{r}
# Cluster the survey variables
hc_survey <- HClustVar$new(vartype = "quant", dist.metric = "rsquare")
hc_survey$fit(survey_data)

# Find optimal clustering
hc_survey$plot_cohesion()
```

```{r}
# Cut into 4 clusters (based on elbow)
hc_survey$cut_tree(k = 4)

# Assess quality
hc_survey$plot_silhouette()
```

```{r}
# Examine results
results <- hc_survey$summary()
print(results$clust_summary)
print(results$clust_members)
```

### Interpretation

```{r}
# Identify what each cluster represents
for (i in 1:4) {
  cat(sprintf("\n=== Cluster %d ===\n", i))
  cluster_vars <- names(hc_survey$labels)[hc_survey$labels == i]
  cat(paste("  -", cluster_vars, collapse = "\n"), "\n")
}
```

**Expected grouping**:
1. **Demographic cluster**: age, income
2. **Product experience cluster**: product_quality, price_satisfaction, delivery_speed
3. **Service cluster**: customer_service, technical_support, response_time
4. **Loyalty cluster**: recommendation_score, repurchase_intent, brand_trust

### Creating Composite Scores

```{r}
# Use centroids as composite scores
composite_scores <- as.data.frame(hc_survey$centroids)
colnames(composite_scores) <- c("Demo_Score", "Product_Score", 
                                 "Service_Score", "Loyalty_Score")

# Add to original data
survey_with_scores <- cbind(survey_data, composite_scores)

# Analyze relationships
cor(composite_scores)
```

```{r}
# Example: Predict loyalty from other factors
loyalty_model <- lm(Loyalty_Score ~ Product_Score + Service_Score, 
                    data = composite_scores)
summary(loyalty_model)
```

---

## Conclusion

### Key Takeaways

1. **HClustVar clusters variables, not observations** - useful for feature reduction and understanding variable relationships

2. **Three variable types supported**: quantitative (correlation-based), qualitative (Cramér's V), and mixed (automatic discretization)

3. **Complete diagnostic toolkit**: cohesion plots, dendrograms, and silhouette analysis help select optimal clusters

4. **Centroids are synthetic variables**: first principal components that summarize each cluster

5. **Prediction capability**: assign new variables to existing clusters without refitting

### When to Use HClustVar

✓ You have many correlated variables  
✓ You need interpretable variable groups  
✓ You want to reduce dimensionality while maintaining variable meanings  
✓ You're dealing with multicollinearity in regression  
✓ You want to select representative variables from groups  

### Further Reading

- **Original paper**: Chavent et al. (2012), "ClustOfVar: An R Package for the Clustering of Variables"
- **Related methods**: PCA, Factor Analysis, Variable selection techniques
- **Package documentation**: `?HClustVar` for detailed method documentation

---

## Appendix: Quick Reference

### Constructor

```{r eval = FALSE}
hc <- HClustVar$new(
  vartype = "auto",          # "quant", "qual", "mixed", "auto"
  dist.metric = "rsquare",   # "r", "rsquare" (for quant only)
  cah.method = "ward.D"      # "ward.D", "single", "complete", etc.
)
```

### Main Methods

```{r eval = FALSE}
hc$fit(data)                      # Fit the model
hc$cut_tree(k = 3)                # Cut into k clusters
hc$cut_tree(h = 1.5)              # Cut at height h
hc$predict(new_data)              # Predict new variables
hc$summary()                      # Detailed statistics
hc$plot_dendrogram(k = 3)         # Dendrogram with clusters
hc$plot_cohesion()                # Cohesion gain curve
hc$plot_silhouette()              # Silhouette plot
print(hc)                         # Model summary
```

### Accessing Results

```{r eval = FALSE}
hc$labels                 # Cluster assignments
hc$n_clusters             # Number of clusters
hc$centroids              # Cluster centroids (matrix)
hc$clusters.eigen         # Eigenvalues per cluster
hc$tree                   # hclust object
hc$dist.matrix            # Distance matrix
hc$vartype                # Detected/specified variable type
hc$dist.metric            # Distance metric used
hc$cah.method             # Clustering method used
```

---

## Session Information

```{r}
sessionInfo()
```
