---
title: "Hierarchical Variable Clustering with HClustVar"
author: "Eugénie Barlet, Modou Mboup, Axel Cano"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Variable Clustering with HClustVar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)
```

# Introduction

This vignette demonstrates the use of hierarchical clustering for variable reduction and grouping. We'll explore three main approaches:

1. **Hierarchical Variable Clustering (HClustVar)** - Groups variables based on similarity
2. **K-Means Clustering (KMeans)** - *[Placeholder for future implementation]*
3. **Quantitative Variable Clustering** - *[Placeholder for future implementation]*

## Loading Required Libraries

```{r libraries}
library(readxl)      # For reading Excel files
library(FactoMineR)  # For factorial analysis
library(R6)          # For R6 class system
library(ggplot2)     # For enhanced visualizations
library(dplyr)       # For data manipulation
library(tidyr)       # For data reshaping
library(knitr)       # For table formatting

# Source your HClustVar class
# source("path/to/HClustVar.R")
```

## Dataset Overview

We'll work with two datasets:

- **Auto2005Non4x4.xls**: Mixed variables (quantitative and qualitative) from automotive data
- **players_22_cleaned.csv**: Quantitative variables from FIFA 22 player statistics

---

# Part 1: Hierarchical Variable Clustering (HClustVar)

## 1.1 Introduction to HClustVar

The `HClustVar` class implements hierarchical clustering of variables using Agglomerative Hierarchical Clustering (AHC). It groups variables based on their similarity, supporting:

- **Quantitative variables**: Using correlation (r) or squared correlation (r²)
- **Qualitative variables**: Using Cramer's V
- **Mixed variables**: Discretization + Cramer's V

### Key Features

- Automatic variable type detection
- Multiple distance metrics and clustering methods
- Comprehensive diagnostics (silhouette, MDS, dendrograms)
- Cluster quality assessment

---

## 1.2 Example 1: Mixed Variables (Automotive Data)

### Loading and Exploring the Data

```{r load-auto-data}
# Load the automotive dataset
auto_data <- read_excel(data("Auto2005Non4x4.xls"))

# Display structure
str(auto_data)

# Summary statistics
summary(auto_data)

# Check variable types
sapply(auto_data, class)
```

### Data Preparation

```{r prepare-auto-data}
# Convert categorical variables to factors
auto_data <- auto_data %>%
  mutate(across(where(is.character), as.factor))

# Check for missing values
colSums(is.na(auto_data))

# Display first rows
head(auto_data) %>% kable()
```

### Fitting HClustVar on Mixed Data

```{r fit-auto-hclust}
# Initialize HClustVar with automatic type detection
hc_auto <- HClustVar$new(
  vartype = "auto",        # Automatic detection
  dist.metric = NULL,      # Will be set automatically
  cah.method = "ward.D"    # Ward's method (recommended)
)

# Fit the model
hc_auto$fit(auto_data)

# Print model summary
print(hc_auto)
```

### Visualizing the Dendrogram

```{r dendrogram-auto}
# Plot dendrogram without cluster highlighting
hc_auto$plot_dendrogram(k = 0)
```

### Determining Optimal Number of Clusters

```{r optimal-k-auto}
# Use the elbow method to find optimal k
hc_auto$plot_agg_levels()

# Try different max_cluster values for better visualization
hc_auto$plot_agg_levels(max_cluster = 15)
```

### Cutting the Tree

```{r cut-tree-auto}
# Cut the tree into 5 clusters (adjust based on elbow plot)
hc_auto$cut_tree(k = 5)

# Visualize dendrogram with cluster colors
hc_auto$plot_dendrogram(k = 5)
```

### Analyzing Clustering Results

```{r summary-auto}
# Get detailed summary
results_auto <- hc_auto$summary()

# Cluster-level summary
cat("\n=== CLUSTER SUMMARY ===\n")
print(results_auto$clust_summary)

# Variable assignments
cat("\n=== VARIABLE ASSIGNMENTS (Top 10) ===\n")
head(results_auto$clust_members, 10) %>% kable()

# Cluster centroid correlations
cat("\n=== CORRELATIONS BETWEEN CLUSTER CENTROIDS ===\n")
print(results_auto$centroids_correlations)

# Overall quality metrics
cat("\n=== OVERALL QUALITY ===\n")
cat(sprintf("Average Silhouette: %.3f\n", results_auto$avg_silhouette))
cat(sprintf("Total Variance Explained: %.2f%%\n", results_auto$total_var_explained))
```

### Quality Assessment

```{r quality-auto}
# Silhouette plot
hc_auto$plot_silhouette()

# Multidimensional scaling projection
hc_auto$mds_projection()
```

### Interpreting the Results

```{r interpret-auto}
# Variables by cluster
cat("\n=== VARIABLES BY CLUSTER ===\n\n")
for (i in 1:hc_auto$n_clusters) {
  cluster_vars <- names(hc_auto$labels)[hc_auto$labels == i]
  cat(sprintf("Cluster %d (%d variables):\n", i, length(cluster_vars)))
  cat(paste("  -", cluster_vars, collapse = "\n"), "\n\n")
}

# Identify best and worst classified variables
cat("\n=== CLASSIFICATION QUALITY ===\n\n")
clust_members <- results_auto$clust_members

# Best classified (low 1 - R2_ratio, high silhouette)
cat("Best classified variables:\n")
best <- clust_members %>%
  arrange(`1 - R2_ratio`) %>%
  head(5)
print(best)

# Worst classified (high 1 - R2_ratio, low silhouette)
cat("\nWorst classified variables:\n")
worst <- clust_members %>%
  arrange(desc(`1 - R2_ratio`)) %>%
  head(5)
print(worst)
```

---

## 1.3 Example 2: Quantitative Variables (FIFA Players)

### Loading and Exploring the Data

```{r load-fifa-data}
# Load the FIFA players dataset
fifa_data <- read.csv("players_22_cleaned.csv", stringsAsFactors = FALSE)

# Display structure
str(fifa_data)

# Select only numerical variables for clustering
fifa_numeric <- fifa_data %>%
  select(where(is.numeric))

# Remove any ID or irrelevant columns if present
# fifa_numeric <- fifa_numeric %>% select(-player_id, -some_other_id)

# Summary statistics
summary(fifa_numeric)

# Check for missing values
colSums(is.na(fifa_numeric))
```

### Data Preparation

```{r prepare-fifa-data}
# Remove columns with too many missing values (>50%)
missing_threshold <- 0.5
fifa_clean <- fifa_numeric[, colMeans(is.na(fifa_numeric)) < missing_threshold]

# Display remaining variables
cat(sprintf("Number of variables: %d\n", ncol(fifa_clean)))
cat(sprintf("Number of observations: %d\n", nrow(fifa_clean)))

# Display first rows
head(fifa_clean) %>% kable()
```

### Fitting HClustVar on Quantitative Data

```{r fit-fifa-hclust}
# Initialize HClustVar for quantitative variables
hc_fifa <- HClustVar$new(
  vartype = "quant",          # Explicit quantitative
  dist.metric = "rsquare",    # Use R² (default)
  cah.method = "ward.D"       # Ward's method
)

# Fit the model
hc_fifa$fit(fifa_clean)

# Print model summary
print(hc_fifa)
```

### Comparing Distance Metrics

```{r compare-metrics-fifa}
# Fit with different metrics for comparison
hc_fifa_r <- HClustVar$new(vartype = "quant", dist.metric = "r", cah.method = "ward.D")
hc_fifa_r$fit(fifa_clean)

# Compare dendrograms
par(mfrow = c(1, 2))
hc_fifa$plot_dendrogram(k = 0)
title(main = "Distance: R²")
hc_fifa_r$plot_dendrogram(k = 0)
title(main = "Distance: |r|")
par(mfrow = c(1, 1))
```

### Determining Optimal Number of Clusters

```{r optimal-k-fifa}
# Elbow method
hc_fifa$plot_agg_levels()

# Zoom on first 20 clusters
hc_fifa$plot_agg_levels(max_cluster = 20)
```

### Cutting the Tree

```{r cut-tree-fifa}
# Cut the tree into optimal clusters (e.g., 8)
hc_fifa$cut_tree(k = 8)

# Visualize dendrogram with clusters
hc_fifa$plot_dendrogram(k = 8)
```

### Analyzing Clustering Results

```{r summary-fifa}
# Get detailed summary
results_fifa <- hc_fifa$summary()

# Cluster-level summary
cat("\n=== CLUSTER SUMMARY ===\n")
results_fifa$clust_summary %>% kable()

# Variable assignments (sorted by cluster and R²)
cat("\n=== VARIABLE ASSIGNMENTS ===\n")
results_fifa$clust_members %>% 
  arrange(cluster, desc(own_cluster_R2)) %>%
  kable()

# Centroid correlations
cat("\n=== CORRELATIONS BETWEEN CLUSTER CENTROIDS ===\n")
results_fifa$centroids_correlations %>% 
  arrange(desc(squared_correlations)) %>%
  head(10) %>%
  kable()

# Overall quality
cat("\n=== OVERALL QUALITY ===\n")
cat(sprintf("Average Silhouette: %.3f\n", results_fifa$avg_silhouette))
cat(sprintf("Total Variance Explained: %.2f%%\n", results_fifa$total_var_explained))
```

### Quality Assessment

```{r quality-fifa}
# Silhouette plot
hc_fifa$plot_silhouette()

# MDS projection
hc_fifa$mds_projection()
```

### Using Clusters for Dimensionality Reduction

```{r dimensionality-reduction-fifa}
# Extract cluster centroids
centroids <- hc_fifa$centroids

# Display centroid dimensions
cat(sprintf("Original variables: %d\n", ncol(fifa_clean)))
cat(sprintf("Cluster centroids: %d\n", ncol(centroids)))
cat(sprintf("Dimensionality reduction: %.1f%%\n", 
            (1 - ncol(centroids)/ncol(fifa_clean)) * 100))

# Use centroids as reduced feature set
fifa_reduced <- as.data.frame(centroids)
colnames(fifa_reduced) <- paste0("Cluster_", 1:ncol(centroids))

head(fifa_reduced) %>% kable()
```

---

## 1.4 Advanced Features

### Prediction on New Variables

```{r prediction-example, eval=FALSE}
# Create some new variables (example)
new_vars <- data.frame(
  new_stat_1 = rnorm(nrow(fifa_clean)),
  new_stat_2 = rnorm(nrow(fifa_clean))
)

# Predict cluster assignment
predictions <- hc_fifa$predict(new_vars)
print(predictions)
```

### Comparing Clustering Methods

```{r compare-methods}
# Compare different hierarchical methods
methods <- c("ward.D", "complete", "average", "single")
silhouette_scores <- numeric(length(methods))

for (i in seq_along(methods)) {
  hc_temp <- HClustVar$new(
    vartype = "quant", 
    dist.metric = "rsquare", 
    cah.method = methods[i]
  )
  hc_temp$fit(fifa_clean)
  hc_temp$cut_tree(k = 8)
  
  results_temp <- hc_temp$summary()
  silhouette_scores[i] <- results_temp$avg_silhouette
}

# Display comparison
comparison <- data.frame(
  Method = methods,
  Avg_Silhouette = round(silhouette_scores, 3)
) %>%
  arrange(desc(Avg_Silhouette))

cat("\n=== METHOD COMPARISON (by Average Silhouette) ===\n")
comparison %>% kable()

# Visualize
ggplot(comparison, aes(x = reorder(Method, Avg_Silhouette), y = Avg_Silhouette)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_text(aes(label = sprintf("%.3f", Avg_Silhouette)), 
            vjust = -0.5, size = 4) +
  coord_cartesian(ylim = c(0, max(silhouette_scores) * 1.1)) +
  labs(
    title = "Clustering Method Comparison",
    x = "Method",
    y = "Average Silhouette Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Custom Visualization: Cluster Heatmap

```{r cluster-heatmap}
# Create a heatmap showing variable-cluster associations
clust_members <- results_fifa$clust_members

# Prepare data for heatmap
heatmap_data <- clust_members %>%
  mutate(variable = rownames(.)) %>%
  select(variable, cluster, own_cluster_R2, next_closest_R2) %>%
  pivot_longer(
    cols = c(own_cluster_R2, next_closest_R2),
    names_to = "metric",
    values_to = "value"
  )

# Plot
ggplot(heatmap_data %>% filter(metric == "own_cluster_R2"), 
       aes(x = factor(cluster), y = variable, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "white", 
    mid = "lightblue", 
    high = "darkblue",
    midpoint = 0.5,
    name = "R²"
  ) +
  labs(
    title = "Variable-Cluster Association Heatmap",
    x = "Cluster",
    y = "Variable"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 7),
    panel.grid = element_blank()
  )
```

---

# Part 2: K-Means Clustering (Placeholder)

*This section will be completed when the KMeans class is implemented.*

## 2.1 Introduction to K-Means

K-Means is a centroid-based clustering algorithm that partitions observations into k clusters.

### Expected Features

- Initialization methods (random, k-means++)
- Multiple distance metrics
- Elbow method and silhouette analysis
- Cluster stability assessment

### Example Structure

```{r kmeans-placeholder, eval=FALSE}
# Initialize KMeans
km <- KMeans$new(
  n_clusters = 5,
  init_method = "kmeans++",
  max_iter = 300
)

# Fit on data
km$fit(fifa_clean)

# Analyze results
km$summary()
km$plot_elbow()
km$plot_clusters()
```

---

# Part 3: Quantitative Variable Clustering (Placeholder)

*This section will be completed when the quantitative variable clustering class is implemented.*

## 3.1 Introduction to Quantitative Variable Clustering

Specialized clustering for quantitative variables using PCA-based approaches.

### Expected Features

- Optimal cluster selection via variance criteria
- Variable rotation and consolidation
- Integration with PCA results

### Example Structure

```{r quant-clust-placeholder, eval=FALSE}
# Initialize Quantitative Variable Clustering
qvc <- QuantVarClust$new(
  method = "pca",
  rotation = "varimax"
)

# Fit on data
qvc$fit(fifa_clean)

# Analyze results
qvc$summary()
qvc$plot_loadings()
```

---

# Comparison of Methods

## When to Use Each Method?

| Method | Best For | Pros | Cons |
|--------|----------|------|------|
| **HClustVar** | Variable reduction, mixed data | Flexible, hierarchical structure | Computationally intensive for large p |
| **K-Means** | Observation clustering, large datasets | Fast, scalable | Requires k specification, sensitive to initialization |
| **Quant Var Clust** | Pure quantitative analysis | Optimal for continuous data | Limited to numeric variables |

## Example Comparison Workflow

```{r comparison-workflow, eval=FALSE}
# Apply all three methods on the same dataset
data <- fifa_clean

# 1. HClustVar
hc <- HClustVar$new(vartype = "quant", dist.metric = "rsquare")
hc$fit(data)
hc$cut_tree(k = 5)

# 2. KMeans (placeholder)
# km <- KMeans$new(n_clusters = 5)
# km$fit(data)

# 3. QuantVarClust (placeholder)
# qvc <- QuantVarClust$new()
# qvc$fit(data)

# Compare results
# comparison_table <- compare_methods(hc, km, qvc)
# print(comparison_table)
```

---

# Practical Recommendations

## Best Practices for HClustVar

### 1. Data Preparation

- **Check variable types**: Ensure factors are properly encoded
- **Handle missing values**: Use appropriate imputation or removal
- **Scale awareness**: HClustVar handles scaling internally for PCA
- **Remove redundant variables**: Check for perfect correlations

### 2. Parameter Selection

- **vartype = "auto"**: Good default for exploratory analysis
- **dist.metric**: 
  - Use `"rsquare"` if you want to ignore correlation direction
  - Use `"r"` if positive vs negative correlation matters
- **cah.method**: `"ward.D"` is recommended for most cases

### 3. Determining k

- Use `plot_agg_levels()` with elbow method
- Consider domain knowledge
- Check silhouette scores for different k values
- Aim for meaningful, interpretable clusters

### 4. Quality Assessment

- **Average silhouette > 0.5**: Good clustering
- **Average silhouette 0.25-0.5**: Reasonable clustering
- **Average silhouette < 0.25**: Consider different k or method
- Check `1 - R2_ratio` for individual variables (low is better)

### 5. Interpretation

- Examine cluster composition
- Look at centroid correlations
- Consider variable importance within clusters (`own_cluster_R2`)
- Use domain knowledge to validate groupings

---

# Troubleshooting

## Common Issues and Solutions

### Issue 1: Error "Tree has not been cut yet"

**Solution**: Call `cut_tree(k)` before `summary()` or `plot_silhouette()`

```{r troubleshoot-1, eval=FALSE}
hc$fit(data)
hc$cut_tree(k = 5)  # Don't forget this step!
hc$summary()
```

### Issue 2: Low silhouette scores

**Solutions**:
- Try different number of clusters
- Change clustering method (`cah.method`)
- Check for outlier variables
- Consider if your variables are truly clusterable

### Issue 3: Unbalanced clusters

**Solution**: This may be natural in your data. Consider:
- Different linkage methods (`complete`, `average`)
- Removing singleton variables before clustering
- Using a different k value

### Issue 4: High centroid correlations

**Solution**: High correlations between clusters might indicate:
- Too many clusters (reduce k)
- Natural multicollinearity in your data
- Need for different distance metric

---

# Performance Considerations

## Computational Complexity

- **Distance matrix**: O(p²n) where p = variables, n = observations
- **Hierarchical clustering**: O(p²log p)
- **Total**: O(p²n + p²log p)

## Memory Usage

- **Distance matrix**: p²/2 × 8 bytes (symmetric)
- **Data storage**: n × p × 8 bytes

## Optimization Tips

```{r performance-tips, eval=FALSE}
# For large datasets
# 1. Reduce variables first (remove highly correlated)
cor_matrix <- cor(data, use = "pairwise.complete.obs")
high_cor <- findCorrelation(cor_matrix, cutoff = 0.95)
data_reduced <- data[, -high_cor]

# 2. Sample observations if n is very large (>10,000)
# Note: This is for variable clustering, so sampling rows is acceptable
set.seed(123)
sample_idx <- sample(nrow(data), min(5000, nrow(data)))
data_sample <- data[sample_idx, ]

# 3. Use caching
results <- hc$summary()  # First call: computed
results <- hc$summary()  # Second call: cached (instant)
```

---

# Conclusion

This vignette demonstrated:

1. **HClustVar basics**: Initialization, fitting, and interpretation
2. **Mixed data handling**: Automatic detection and Cramer's V
3. **Quantitative data**: Correlation-based clustering
4. **Quality assessment**: Silhouette, MDS, dendrograms
5. **Advanced features**: Prediction, method comparison, custom visualizations

## Next Steps

- Explore your own datasets with HClustVar
- Experiment with different parameters
- Integrate with downstream analysis (regression, classification)
- Compare with other dimensionality reduction techniques (PCA, FA)

## References

- Vigneau, E., & Qannari, E. M. (2003). Clustering of variables around latent components. *Communications in Statistics - Simulation and Computation*, 32(4), 1131-1150.
- Ward, J. H. (1963). Hierarchical grouping to optimize an objective function. *Journal of the American Statistical Association*, 58(301), 236-244.

---

# Session Info

```{r session-info}
sessionInfo()
```
