---
title: "K-means Variable Clustering with KmeansVariables"
author: "Eugénie Barlet, Modou Mboup, Axel Cano"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Variable Clustering with KmeansVariables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)
```

# Introduction

This vignette demonstrates the use of K-means clustering for variable grouping. The `KmeansVariables` class implements a reallocation-based algorithm that groups variables by maximizing their correlation with cluster centroids computed via Principal Component Analysis (PCA).

## Key Features

- **Iterative reallocation** of variables to clusters
- **Centroid computation via PCA** (first principal component)
- **Multiple random initializations** to avoid local minima
- **Two distance metrics**: R² (absolute correlation) or signed r
- **Automatic K selection** via elbow, silhouette, or Calinski-Harabasz criteria
- **Comprehensive diagnostics**: silhouette, contributions, projections

For hierarchical variable clustering (including qualitative variables), see the dedicated vignette `Variable_clustering.Rmd`.

## Loading Required Libraries

```{r libraries}
library(R6)          # For R6 class system
library(ggplot2)     # For enhanced visualizations
library(dplyr)       # For data manipulation
library(tidyr)       # For data reshaping
library(knitr)       # For table formatting

# Load the package
# library(sise.var.clustering)
# Or source the files directly
# source("R/KmeansVariables.R")
# source("R/metrics_kmeans.R")
# source("R/plot_kmeans.R")
```

## Dataset Overview

We'll work with the **FIFA 22 players dataset** containing 41 quantitative skill variables for 17,107 players.

```{r load-data}
# Load the FIFA players dataset
data("players")

# Display structure
cat(sprintf("Number of observations: %d\n", nrow(players)))
cat(sprintf("Number of variables: %d\n", ncol(players)))

# Display first variables
head(names(players), 20)
```

---

# Part 1: Basic Usage

## 1.1 Introduction to K-means Variable Clustering

The K-means for variables algorithm works as follows:

1. **Initialization**: Randomly assign variables to K clusters
2. **Centroid Calculation**: For each cluster, compute the first principal component (via PCA) as the cluster centroid
3. **Distance Calculation**: Compute distance between each variable and each centroid based on their correlation
4. **Reallocation**: Reassign each variable to its nearest centroid
5. **Iteration**: Repeat steps 2-4 until convergence

### Distance Metrics

Two distance metrics are available:

- **R² (r_squared)**: Groups variables with high absolute correlation (positive OR negative)
  $$d(x_j, c_k) = \sqrt{1 - r^2(x_j, c_k)}$$

- **Signed r (r_signed)**: Groups only variables with high positive correlation
  $$d(x_j, c_k) = 1 - r(x_j, c_k)$$

### Inertia Criterion

The algorithm maximizes within-cluster cohesion (inertia). For cluster $k$, the inertia equals the first eigenvalue $\lambda_k$ from PCA:

$$I_k = \lambda_k = \sum_{j \in C_k} r^2(x_j, c_k)$$

---

## 1.2 Data Preparation

```{r prepare-data}
# Check for missing values
missing_counts <- colSums(is.na(players))
cat("Variables with missing values:", sum(missing_counts > 0), "\n")

# Summary statistics
summary(players[, 1:6])

# Display first rows
head(players) %>% kable()
```

---

## 1.3 Fitting the Model

### Basic Fitting

```{r fit-basic}
# Initialize KmeansVariables with 5 clusters
km <- KmeansVariables$new(
  n_clusters = 5,
  n_init = 10,           # Number of random initializations
  max_iter = 100,        # Maximum iterations per init
  random_state = 42      # For reproducibility
)

# Fit the model
km$fit(players)

# Print model summary
print(km)
```

### Accessing Results

```{r access-results}
# Cluster assignments
cat("Cluster assignments:\n")
head(km$labels, 10)

# Cluster sizes
cat("\nCluster sizes:\n")
km$cluster_sizes

# Total inertia
cat("\nTotal inertia:", round(km$inertia, 4), "\n")

# Inertia per cluster
cat("\nInertia per cluster:\n")
km$cluster_inertias
```

---

## 1.4 Visualizing Results

### Elbow Plot for Optimal K

```{r elbow-plot}
# Plot elbow curve to determine optimal K
km$plot_elbow(k_range = 2:10)
```

The elbow plot shows how total inertia increases with K. Look for the "elbow" point where adding more clusters provides diminishing returns.

### Variable Projection (PCA Circle)

```{r projection-plot}
# Plot variable projection on first two principal components
km$plot_projection(show_labels = TRUE)
```

This plot shows variables grouped by cluster in a PCA-like projection. Variables close together are highly correlated.

### Top Contributing Variables

```{r contributions-plot}
# Plot top 5 contributing variables per cluster
km$plot_contributions(top_n = 5)
```

This plot identifies the most representative variables for each cluster based on their R² with the centroid.

---

## 1.5 Analyzing Results

### Comprehensive Summary

```{r summary-results}
# Get detailed summary
results <- km$summary(print_summary = TRUE)
```

### Cluster Summary

```{r cluster-summary}
# Cluster-level statistics
cat("\n=== CLUSTER SUMMARY ===\n")
results$clust_summary %>% kable()
```

**Interpretation**:
- `n_members`: Number of variables in each cluster
- `variation_explained`: Total variance explained by the cluster (eigenvalue)
- `proportion_explained`: Average variance per variable (quality indicator)

### Variable Assignments

```{r variable-assignments}
# Variable details
cat("\n=== VARIABLE ASSIGNMENTS ===\n")
results$clust_members %>%
  arrange(cluster, desc(own_cluster_R2)) %>%
  kable()
```

**Interpretation**:
- `own_cluster_R2`: Correlation² with own cluster centroid (higher = better fit)
- `next_closest_R2`: Correlation² with second closest centroid
- `1 - R2_ratio`: Adhesion to own cluster vs next closest (lower = better separation)

### Top Variables per Cluster

```{r top-variables}
# Top contributing variables
cat("\n=== TOP VARIABLES PER CLUSTER ===\n")
results$top_variables %>% kable()
```

### Centroid Correlations

```{r centroid-correlations}
# Correlations between cluster centroids
cat("\n=== CORRELATIONS BETWEEN CLUSTER CENTROIDS ===\n")
round(results$centroids_correlations, 3) %>% kable()
```

Low correlations between centroids indicate well-separated clusters.

### Overall Quality Metrics

```{r quality-metrics}
# Quality metrics
cat("\n=== OVERALL QUALITY ===\n")
cat(sprintf("Average Silhouette: %.3f\n", results$avg_silhouette))
cat(sprintf("Total Variance Explained: %.2f%%\n", results$total_var_explained * 100))
```

**Interpretation**:
- **Average Silhouette > 0.5**: Good clustering
- **Average Silhouette 0.25-0.5**: Reasonable clustering
- **Average Silhouette < 0.25**: Consider different K or method
- **Total Variance Explained**: Percentage of total variance captured by clustering

---

## 1.6 Silhouette Analysis

```{r silhouette-analysis}
# Get silhouette coefficients
sil <- km$silhouette()

# Display silhouette values
cat("=== SILHOUETTE COEFFICIENTS ===\n")
sil %>%
  arrange(cluster, desc(silhouette)) %>%
  kable()

# Summary by cluster
cat("\n=== AVERAGE SILHOUETTE BY CLUSTER ===\n")
sil %>%
  group_by(cluster) %>%
  summarise(
    n_variables = n(),
    avg_silhouette = round(mean(silhouette), 3),
    min_silhouette = round(min(silhouette), 3),
    max_silhouette = round(max(silhouette), 3)
  ) %>%
  kable()
```

---

## 1.7 Variable Contributions

```{r contributions-table}
# Get contributions table
contrib <- km$contributions()

# Display contributions
cat("=== VARIABLE CONTRIBUTIONS ===\n")
contrib %>%
  arrange(cluster, desc(contribution)) %>%
  kable()
```

---

## 1.8 Correlation Table

```{r correlation-table}
# Get correlation matrix between variables and centroids
cor_table <- km$correlation_table(round_digits = 3)

# Display correlation table
cat("=== VARIABLE-CENTROID CORRELATIONS ===\n")
cor_table %>% kable()
```

---

# Part 2: Advanced Features

## 2.1 Distance Metrics Comparison

### R² Metric (Default)

```{r rsquare-metric}
# R²-based clustering (groups variables with high |correlation|)
km_rsq <- KmeansVariables$new(
  n_clusters = 5,
  distance_metric = "r_squared",
  random_state = 42
)
km_rsq$fit(players)

cat("R² metric - Total inertia:", round(km_rsq$inertia, 4), "\n")
```

### Signed r Metric

```{r signed-metric}
# Signed r clustering (groups only positively correlated variables)
km_signed <- KmeansVariables$new(
  n_clusters = 5,
  distance_metric = "r_signed",
  random_state = 42
)
km_signed$fit(players)

cat("Signed r metric - Total inertia:", round(km_signed$inertia, 4), "\n")
```

### Comparison

```{r compare-metrics}
# Compare silhouette scores
results_rsq <- km_rsq$summary()
results_signed <- km_signed$summary()

comparison <- data.frame(
  Metric = c("R² (r_squared)", "Signed r (r_signed)"),
  Total_Inertia = c(km_rsq$inertia, km_signed$inertia),
  Avg_Silhouette = c(results_rsq$avg_silhouette, results_signed$avg_silhouette),
  Var_Explained = c(results_rsq$total_var_explained, results_signed$total_var_explained)
)

comparison %>% kable()
```

---

## 2.2 Automatic K Selection

The `KmeansVariables` class supports automatic selection of the optimal number of clusters.

### Using Silhouette Method

```{r auto-silhouette}
# Automatic K selection using silhouette
km_auto_sil <- KmeansVariables$new(
  n_clusters = "auto",
  k_range = 2:10,
  selection_method = "silhouette",
  random_state = 42
)

km_auto_sil$fit(players)

cat("Optimal K (silhouette):", km_auto_sil$n_clusters, "\n")
cat("Inertia:", round(km_auto_sil$inertia, 4), "\n")
```

### Using Elbow Method

```{r auto-elbow}
# Automatic K selection using elbow
km_auto_elbow <- KmeansVariables$new(
  n_clusters = "auto",
  k_range = 2:10,
  selection_method = "elbow",
  random_state = 42
)

km_auto_elbow$fit(players)

cat("Optimal K (elbow):", km_auto_elbow$n_clusters, "\n")
```

### Using Calinski-Harabasz Index

```{r auto-calinski}
# Automatic K selection using Calinski-Harabasz
km_auto_ch <- KmeansVariables$new(
  n_clusters = "auto",
  k_range = 2:10,
  selection_method = "calinski",
  random_state = 42
)

km_auto_ch$fit(players)

cat("Optimal K (Calinski-Harabasz):", km_auto_ch$n_clusters, "\n")
```

### Using All Methods (Consensus)

```{r auto-all}
# Automatic K selection using all methods
km_auto_all <- KmeansVariables$new(
  n_clusters = "auto",
  k_range = 2:10,
  selection_method = "all",
  random_state = 42
)

km_auto_all$fit(players)

cat("Optimal K (consensus):", km_auto_all$n_clusters, "\n")
```

---

## 2.3 Prediction on New Variables

```{r prediction}
# Create some new variables (example with noise)
set.seed(123)
new_vars <- data.frame(
  new_stat_1 = players$pace + rnorm(nrow(players), 0, 5),
  new_stat_2 = players$defending + rnorm(nrow(players), 0, 5),
  new_stat_3 = rnorm(nrow(players))  # Random noise
)

# Predict cluster assignment
predictions <- km$predict(new_vars)

cat("Predictions for new variables:\n")
print(predictions)
```

### Project Supplementary Variables

```{r supplementary-projection}
# Project new variables onto PCA plot
km$plot_projection(
  supplementary_variables = new_vars,
  show_labels = TRUE
)
```

---

## 2.4 Multiple Initializations Impact

```{r multiple-inits}
# Compare results with different number of initializations
inertias <- numeric(5)
init_values <- c(1, 5, 10, 20, 50)

for (i in seq_along(init_values)) {
  km_temp <- KmeansVariables$new(
    n_clusters = 5,
    n_init = init_values[i],
    random_state = 42
  )
  km_temp$fit(players)
  inertias[i] <- km_temp$inertia
}

# Display comparison
init_comparison <- data.frame(
  n_init = init_values,
  inertia = round(inertias, 4)
)

cat("=== IMPACT OF MULTIPLE INITIALIZATIONS ===\n")
init_comparison %>% kable()

# Plot
ggplot(init_comparison, aes(x = n_init, y = inertia)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue", size = 3) +
  labs(
    title = "Impact of Number of Initializations on Inertia",
    x = "Number of Initializations",
    y = "Total Inertia"
  ) +
  theme_minimal()
```

---

## 2.5 Reproducibility with Random State

```{r reproducibility}
# Same random_state = same results
km1 <- KmeansVariables$new(n_clusters = 5, random_state = 123)
km1$fit(players)

km2 <- KmeansVariables$new(n_clusters = 5, random_state = 123)
km2$fit(players)

cat("Same random state produces identical results:\n")
cat("Inertia 1:", km1$inertia, "\n")
cat("Inertia 2:", km2$inertia, "\n")
cat("Labels identical:", all(km1$labels == km2$labels), "\n")
```

---

# Part 3: Interpreting FIFA Player Skill Clusters

## 3.1 Understanding the Clusters

```{r interpret-clusters}
# Get final results
results <- km$summary()

# Display variables by cluster
cat("=== VARIABLES BY CLUSTER ===\n\n")
for (i in 1:km$n_clusters) {
  cluster_vars <- names(km$labels)[km$labels == i]
  cat(sprintf("Cluster %d (%d variables):\n", i, length(cluster_vars)))
  cat(paste("  -", cluster_vars, collapse = "\n"), "\n\n")
}
```

## 3.2 Cluster Characterization

```{r characterize-clusters}
# Identify cluster themes based on variable names
cat("=== CLUSTER THEMES ===\n\n")

for (i in 1:km$n_clusters) {
  cluster_vars <- names(km$labels)[km$labels == i]

  # Count variable prefixes
  prefixes <- sapply(strsplit(cluster_vars, "_"), `[`, 1)
  prefix_table <- sort(table(prefixes), decreasing = TRUE)

  cat(sprintf("Cluster %d:\n", i))
  cat("  Main themes:", paste(names(prefix_table), collapse = ", "), "\n")
  cat("  Top variables:\n")

  # Get top 3 by R²
  top_vars <- results$top_variables %>%
    filter(cluster_id == i) %>%
    arrange(desc(r_squared)) %>%
    head(3)

  for (j in 1:nrow(top_vars)) {
    cat(sprintf("    - %s (R² = %.3f)\n",
                top_vars$variable[j],
                top_vars$r_squared[j]))
  }
  cat("\n")
}
```

## 3.3 Quality Assessment

```{r quality-assessment}
# Identify best and worst classified variables
clust_members <- results$clust_members

# Best classified (high silhouette, low 1 - R2_ratio)
cat("=== BEST CLASSIFIED VARIABLES ===\n")
best <- clust_members %>%
  arrange(`1 - R2_ratio`) %>%
  head(5)
print(best)

# Worst classified (low silhouette, high 1 - R2_ratio)
cat("\n=== WORST CLASSIFIED VARIABLES ===\n")
worst <- clust_members %>%
  arrange(desc(`1 - R2_ratio`)) %>%
  head(5)
print(worst)
```

---

# Part 4: Using Clusters for Dimensionality Reduction

## 4.1 Extract Cluster Centroids

```{r dimensionality-reduction}
# Get centroids (first principal component of each cluster)
centroids <- km$centroids

# Display dimensions
cat("Original variables:", ncol(players), "\n")
cat("Cluster centroids:", ncol(centroids), "\n")
cat("Dimensionality reduction:",
    sprintf("%.1f%%", (1 - ncol(centroids)/ncol(players)) * 100), "\n")
```

## 4.2 Use Centroids as Features

```{r centroids-as-features}
# Use centroids as reduced feature set
players_reduced <- as.data.frame(centroids)
colnames(players_reduced) <- paste0("Cluster_", 1:ncol(centroids))

# Display reduced data
cat("\n=== REDUCED DATASET (First 6 rows) ===\n")
head(players_reduced) %>% kable()

# Correlation between new features
cat("\n=== CORRELATION BETWEEN CLUSTER FEATURES ===\n")
round(cor(players_reduced), 3) %>% kable()
```

---

# Practical Recommendations

## Best Practices for KmeansVariables

### 1. Data Preparation

- **Check variable types**: Ensure all variables are numeric
- **Handle missing values**: The algorithm handles them, but consider imputation
- **Remove constant variables**: Variables with zero variance cause issues
- **Scale awareness**: PCA handles scaling internally

### 2. Parameter Selection

- **n_clusters**: Use elbow plot, silhouette, or domain knowledge
- **n_init**: At least 10 for stable results (default)
- **distance_metric**:
  - Use `"r_squared"` if you want to group variables with high |correlation|
  - Use `"r_signed"` if only positive correlation should matter
- **random_state**: Set for reproducibility

### 3. Determining K

- Use `plot_elbow()` for visual inspection
- Use `n_clusters = "auto"` with different `selection_method` values
- Consider domain knowledge and interpretability
- Try multiple K values and compare silhouette scores

### 4. Quality Assessment

- **Average silhouette > 0.5**: Good clustering
- **Average silhouette 0.25-0.5**: Reasonable clustering
- **Average silhouette < 0.25**: Consider different K
- Check `1 - R2_ratio` for individual variables (low is better)
- Low centroid correlations indicate good separation

### 5. Interpretation

- Examine variable names in each cluster
- Look for thematic groupings
- Use `plot_contributions()` to identify key variables
- Validate with domain expertise

---

# Troubleshooting

## Common Issues and Solutions

### Issue 1: Low inertia or silhouette

**Solutions**:
- Try different number of clusters
- Increase `n_init` for more initializations
- Check for outlier variables
- Try different distance metric

### Issue 2: Unstable results

**Solution**: Increase `n_init` or set `random_state`

```{r troubleshoot-unstable, eval=FALSE}
km <- KmeansVariables$new(
  n_clusters = 5,
  n_init = 50,       # More initializations
  random_state = 42  # For reproducibility
)
```

### Issue 3: Empty clusters

**Solution**: The algorithm handles this automatically by reassigning variables

### Issue 4: High centroid correlations

**Solution**: Reduce number of clusters or use different distance metric

---

# Comparison: K-means vs Hierarchical Clustering

| Aspect | KmeansVariables | HClustVar |
|--------|-----------------|-----------|
| **Algorithm** | Reallocation-based | Agglomerative |
| **Output** | Flat partition | Dendrogram + cut |
| **K selection** | Must specify or auto | Cut at any level |
| **Speed** | Fast | Slower for large p |
| **Determinism** | Depends on init | Deterministic |
| **Variable types** | Quantitative only | Mixed supported |

---

# Conclusion

This vignette demonstrated:

1. **Basic usage**: Fitting, visualization, and interpretation
2. **Distance metrics**: R² vs signed r comparison
3. **Automatic K selection**: Multiple criteria available
4. **Prediction**: Assigning new variables to clusters
5. **Dimensionality reduction**: Using centroids as features
6. **Quality assessment**: Silhouette, contributions, correlations

## Next Steps

- Compare results with `HClustVar` on the same data
- Use cluster centroids in downstream analysis
- Experiment with different K values and distance metrics
- Apply to your own datasets

## References

- Vigneau, E., & Qannari, E. M. (2003). Clustering of variables around latent components. *Communications in Statistics - Simulation and Computation*, 32(4), 1131-1150.
- Chavent, M., et al. (2012). ClustOfVar: An R Package for the Clustering of Variables. *Journal of Statistical Software*, 50(13).

---

# Session Info

```{r session-info}
sessionInfo()
```
