---
title: "Modality Clustering with ModClust"
author: "Eugénie Barlet, Modou Mboup, Axel Cano"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Modality Clustering with ModClust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)
```

# Introduction

This vignette demonstrates the use of **Modality-Based Hierarchical Clustering (ModClust)** for grouping modalities (categories) of categorical variables based on their co-occurrence patterns.

ModClust uses:
- **Dice distance** for measuring dissimilarity between modalities
- **Hierarchical clustering** with various linkage methods
- **Multiple Correspondence Analysis (MCA)** for visualization

For variable clustering (quantitative or mixed), see the dedicated vignettes `KmeansVariables.Rmd` and `Variable_clustering.Rmd`.

## Loading the library

```{r libraries}
library(VarClustering)
```

## Dataset Overview

We'll work with the **Congressional Voting Records** dataset containing 435 observations and 7 categorical variables representing votes on various issues.

---

# Part 1: Modality-Based Clustering (ModClust)

## 1.1 Introduction to ModClust

The `ModClust` class implements hierarchical clustering of modalities using the Dice distance coefficient. It clusters the categories (modalities) of categorical variables rather than observations.

### Algorithm Overview

1. **Complete Disjunctive Table**: Convert categorical data to binary indicators
2. **Dice Distance**: Compute dissimilarity between all pairs of modalities
3. **Hierarchical Clustering**: Apply agglomerative clustering
4. **MCA Visualization**: Project modalities in reduced dimensional space

### Dice Distance Formula

The Dice distance between modalities j and j' is:

$$D(j, j') = \sqrt{\frac{1}{2} \sum_{i=1}^{n} (x_{ij}(1-x_{ij'}) + (1-x_{ij})x_{ij'})}$$

where $x_{ij}$ is the indicator (0 or 1) that individual i has modality j.

### Key Features

- Specialized for categorical data clustering
- Multiple linkage methods (Ward, average, complete, single)
- MCA integration for visualization
- Prediction for new observations
- Quality assessment with silhouette coefficients

---

## 1.2 Example: Congressional Voting Dataset

### Loading and Exploring the Data

```{r load-vote}
# Load the vote dataset
data(vote)

# Display structure
cat("Dimensions:", dim(vote), "\n")
cat("Variables:", paste(names(vote), collapse = ", "), "\n")

# Summary
summary(vote)
```

### Data Exploration

```{r explore-vote}
# Check variable types
sapply(vote, class)

# Display first rows
head(vote)

# Count modalities per variable
sapply(vote, function(x) length(levels(x)))
```

### Splitting Data for Prediction Example

```{r split-data}
# Split into training and test sets
set.seed(42)
train_idx <- sample(nrow(vote), size = round(0.8 * nrow(vote)))
vote_train <- vote[train_idx, ]
vote_test <- vote[-train_idx, ]

cat("Training set:", nrow(vote_train), "observations\n")
cat("Test set:", nrow(vote_test), "observations\n")
```

---

## 1.3 Fitting the Model

### Basic Fitting

```{r fit-modclust}
# Initialize ModClust
mc <- ModClust$new(
  n_clusters = NULL,       # Will be determined later
  method = "hclust",       # Hierarchical clustering
  n_dimensions = 5,        # MCA dimensions for visualization
  hclust_method = "ward.D2" # Ward's method (recommended)
)

# Fit the model on training data
mc$fit(vote_train)

# Print model summary
print(mc)
```

### Accessing Basic Results

```{r access-results}
# Modality names
cat("Number of modalities:", length(mc$modality_names), "\n")
cat("Modality names (first 10):\n")
head(mc$modality_names, 10)

# Modality frequencies
cat("\nModality frequencies (first 10):\n")
head(mc$modality_frequencies, 10)
```

---

## 1.4 Determining Optimal Number of Clusters

### Aggregation Heights Plot (Elbow Method)

```{r plot-heights}
# Plot aggregation heights to find optimal k
mc$plot_heights()
```

Look for the "elbow" point where the aggregation height increases sharply, indicating the optimal number of clusters.

### Cutting the Tree

```{r cut-tree}
# Cut the tree into 4 clusters (adjust based on elbow plot)
mc$cut_tree(k = 4)

# Print updated summary
print(mc)
```

---

## 1.5 Visualizing Results

### Dendrogram

```{r dendrogram}
# Plot dendrogram with colored branches
mc$plot_dendrogram(k = 4, color_branches = TRUE)
```

### MCA Projection

```{r mca-plot}
# Plot modalities in MCA space with cluster colors
mc$plot_mca(axes = c(1, 2), show_labels = TRUE, label_cex = 0.7)
```

The MCA plot shows modalities positioned according to their co-occurrence patterns. Modalities that appear together in observations are closer in the plot.

### Silhouette Plot

```{r silhouette}
# Plot silhouette coefficients for quality assessment
mc$plot_silhouette()
```

Silhouette values range from -1 to 1. High values indicate good cluster assignment.

---

## 1.6 Analyzing Results

### Comprehensive Summary

```{r summary}
# Get detailed summary
results <- mc$summary()

# Display cluster summary
cat("=== CLUSTER SUMMARY ===\n")
results$clust_summary
```

### Cluster Members Details

```{r clust-members}
# Display modality-level details
cat("\n=== CLUSTER MEMBERS ===\n")
results$clust_members
```

**Interpretation**:
- `frequency`: Number of occurrences of the modality in the data
- `dist_own`: Distance to own cluster centroid (lower = better fit)
- `dist_nearest`: Distance to nearest other cluster
- `ratio`: Ratio of distances (lower = better separation)

### Cluster Table

```{r cluster-table}
# Simple table of modality-cluster assignments
cluster_table <- mc$get_cluster_table()
cluster_table
```

### Modalities by Cluster

```{r modalities-by-cluster}
# Display modalities grouped by cluster
cat("=== MODALITIES BY CLUSTER ===\n\n")
for (i in 1:4) {
  cluster_mods <- cluster_table$Modality[cluster_table$Cluster == i]
  cat(sprintf("Cluster %d (%d modalities):\n", i, length(cluster_mods)))
  cat(paste("  -", cluster_mods, collapse = "\n"), "\n\n")
}
```

---

## 1.7 Prediction on New Observations

### Predicting Cluster Membership

```{r predict}
# Predict cluster for new observations
predictions <- mc$predict(vote_test)

# Display predictions (first 10)
cat("=== PREDICTIONS (first 10) ===\n")
head(predictions, 10)
```

The prediction assigns each observation to the cluster whose centroid has the smallest Dice distance to the observation's modality profile.

### Projection of New Observations

```{r project-new}
# Project new observations into MCA space
new_coords <- mc$project_new_modalities(vote_test, axes = c(1, 2))

# Display coordinates (first 5)
cat("\n=== PROJECTED COORDINATES (first 5) ===\n")
head(new_coords, 5)
```

### Visualizing Training and New Observations

```{r plot-with-new}
# Plot both training modalities and new observations
mc$plot_mca_with_new(
  new_data = vote_test,
  axes = c(1, 2),
  show_labels = TRUE,
  label_cex = 0.6
)
```

Training modalities appear as points, new observations as stars.

---

## 1.8 Dice Distance Matrix

```{r dice-matrix}
# Get the complete Dice distance matrix
dice_matrix <- mc$get_dice_matrix()

# Display a subset (first 5x5)
cat("=== DICE DISTANCE MATRIX (subset) ===\n")
round(dice_matrix[1:5, 1:5], 3)
```

---

# Part 2: Advanced Features

## 2.1 Comparing Linkage Methods

```{r compare-methods}
# Compare different hierarchical clustering methods
methods <- c("ward.D2", "average", "complete", "single")
results_list <- list()

for (m in methods) {
  mc_temp <- ModClust$new(
    n_clusters = 4,
    hclust_method = m,
    n_dimensions = 5
  )
  mc_temp$fit(vote_train)

  # Get summary
  res <- mc_temp$summary()

  # Store cluster sizes
  results_list[[m]] <- data.frame(
    Method = m,
    n_modalities = nrow(res$clust_members),
    avg_dist_own = mean(res$clust_members$dist_own),
    avg_ratio = mean(res$clust_members$ratio)
  )
}

# Combine results
comparison <- do.call(rbind, results_list)
cat("=== METHOD COMPARISON ===\n")
comparison
```

## 2.2 Different Number of Clusters

```{r different-k}
# Test different numbers of clusters
k_values <- 2:6

for (k in k_values) {
  mc$cut_tree(k = k)
  res <- mc$summary()

  # Calculate average ratio (quality metric)
  avg_ratio <- mean(res$clust_members$ratio, na.rm = TRUE)
  cat(sprintf("K = %d: %d clusters, avg ratio = %.3f\n",
              k, k, avg_ratio))
}

# Reset to optimal k
mc$cut_tree(k = 4)
```

## 2.3 Interpreting Cluster Themes

```{r interpret-clusters}
# Analyze cluster composition
cluster_table <- mc$get_cluster_table()

cat("=== CLUSTER INTERPRETATION ===\n\n")

for (i in 1:4) {
  cluster_mods <- cluster_table$Modality[cluster_table$Cluster == i]
  cluster_freq <- cluster_table$Frequency[cluster_table$Cluster == i]

  # Extract variable names from modality names
  vars <- sapply(strsplit(cluster_mods, "\\."), `[`, 1)
  var_counts <- sort(table(vars), decreasing = TRUE)

  cat(sprintf("Cluster %d:\n", i))
  cat("  Main variables:", paste(names(var_counts), collapse = ", "), "\n")
  cat("  Top modalities by frequency:\n")

  # Show top 3 by frequency
  top_idx <- order(cluster_freq, decreasing = TRUE)[1:min(3, length(cluster_freq))]
  for (j in top_idx) {
    cat(sprintf("    - %s (freq: %d)\n", cluster_mods[j], cluster_freq[j]))
  }
  cat("\n")
}
```

---

# Practical Recommendations

## Best Practices for ModClust

### 1. Data Preparation

- **All variables must be factors**: Convert character columns to factors
- **Handle missing values**: ModClust may not handle NA values
- **Check modality counts**: Very rare modalities may cluster poorly

### 2. Parameter Selection

- **hclust_method**: `"ward.D2"` is recommended for most cases
- **n_dimensions**: Higher values (5-10) for complex data
- **n_clusters**: Use `plot_heights()` elbow method

### 3. Determining k

- Use `plot_heights()` for elbow detection
- Consider interpretability of clusters
- Check silhouette values for different k values

### 4. Quality Assessment

- **Silhouette > 0.5**: Good clustering
- **Silhouette 0.25-0.5**: Reasonable clustering
- **Silhouette < 0.25**: Consider different k or method
- Check `ratio` column in summary (lower = better separation)

### 5. Interpretation

- Examine modality names in each cluster
- Look for thematic patterns (same variable, similar values)
- Use MCA plot to visualize proximity
- Validate with domain expertise

---

# Troubleshooting

## Common Issues and Solutions

### Issue 1: Error with non-factor variables

**Solution**: Convert all variables to factors

```{r troubleshoot-factors, eval=FALSE}
vote <- vote %>%
  mutate(across(everything(), as.factor))
```

### Issue 2: Low silhouette scores

**Solutions**:
- Try different number of clusters
- Change linkage method
- Check for problematic modalities (very rare)

### Issue 3: Unbalanced clusters

**Solution**: This may be natural. Consider:
- Different linkage methods
- Removing very rare modalities
- Different k value

### Issue 4: Prediction errors

**Solution**: Ensure new data has same factor levels as training

```{r troubleshoot-levels, eval=FALSE}
# Check levels match
all(levels(new_data$var) %in% levels(train_data$var))
```

---

# Comparison: ModClust vs Variable Clustering

| Aspect | ModClust | HClustVar / KmeansVariables |
|--------|----------|---------------------------|
| **Clusters** | Modalities (categories) | Variables |
| **Data type** | Categorical only | Quantitative / Mixed |
| **Distance** | Dice coefficient | Correlation / Cramer's V |
| **Visualization** | MCA | PCA / MDS |
| **Use case** | Find patterns in categorical data | Reduce variable dimensionality |

---

# Conclusion

This vignette demonstrated **ModClust** for modality-based hierarchical clustering:

- Dice distance for categorical data dissimilarity
- Hierarchical clustering with various linkage methods
- MCA visualization of modality clusters
- Prediction and projection of new observations
- Quality assessment with silhouette coefficients

For variable clustering, see the dedicated vignettes:
- `Variable_clustering.Rmd` for hierarchical variable clustering (HClustVar)
- `KmeansVariables.Rmd` for K-means variable clustering

---

# Session Info

```{r session-info}
sessionInfo()
```
