---
title: "Variable and Modality Clustering Package"
author: "Eugénie Barlet, Modou Mboup, Axel Cano"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Variable and Modality Clustering Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)
```

# Introduction

This vignette provides a comprehensive guide to the **VarClustering** package, which implements three complementary clustering approaches for variable reduction and pattern discovery:

1. **HClustVar** - Hierarchical clustering of variables (quantitative, qualitative, or mixed)
2. **KmeansVariables** - K-means clustering of quantitative variables using PCA centroids
3. **ModCluster** - Modality-based clustering using Dice distance for categorical data

## Package Overview

| Algorithm | Data Type | Distance/Similarity | Output |
|-----------|-----------|---------------------|--------|
| HClustVar | Quantitative / Qualitative / Mixed | R, R², Cramer's V | Variable clusters |
| KmeansVariables | Quantitative only | R, R² | Variable clusters |
| ModCluster | Categorical only | Dice distance | Modality clusters |

## Loading the Package

```{r libraries}
library(VarClustering)
```

## Datasets

We use four datasets throughout this vignette:

- **autos**: Quantitative automotive variables
- **canines**: Qualitative dog characteristics
- **players**: FIFA 22 player skill statistics (quantitative)
- **vote**: Congressional voting records (categorical)

---

# Part 1: Hierarchical Variable Clustering (HClustVar)

## 1.1 Introduction

The `HClustVar` class implements hierarchical clustering of variables using Agglomerative Hierarchical Clustering (AHC). It groups variables based on their similarity, supporting:

- **Quantitative variables**: Correlation (r) or squared correlation (r²)
- **Qualitative variables**: Cramer's V
- **Mixed variables**: Discretization + Cramer's V

### Key Features

- Automatic variable type detection
- Multiple distance metrics and linkage methods
- Comprehensive diagnostics (silhouette, MDS, dendrograms)
- Prediction on illustrative variables

---

## 1.2 Example 1: Quantitative Variables (Automotive Data)

### Loading and Exploring

```{r hclust-load-auto}
# Load the autos dataset
data(autos)

# Display structure
str(autos)

# Summary
summary(autos)
```

### Splitting Data

```{r hclust-split-auto}
# Split into active and illustrative variables
autos_train <- autos[, 2:9]
autos_test <- autos[, 10:12]
```

### Fitting the Model

```{r hclust-fit-auto}
# Initialize HClustVar
hc_auto <- HClustVar$new(
  vartype = "quant",
  dist.metric = "rsquare",
  cah.method = "ward.D2"
)

# Fit the model
hc_auto$fit(autos_train)

# Print summary
print(hc_auto)
```

### Visualizing the Dendrogram

```{r hclust-dendro-auto}
hc_auto$plot_dendrogram()
```

### Determining Optimal K

```{r hclust-elbow-auto}
hc_auto$plot_agg_levels()
```

### Cutting the Tree

```{r hclust-cut-auto}
# Cut into 3 clusters
hc_auto$cut_tree(k = 3)

# Visualize with cluster colors
hc_auto$plot_dendrogram(k = 3)
```

### Quality Assessment

```{r hclust-quality-auto}
# MDS projection
hc_auto$mds_projection()

# Silhouette plot
hc_auto$plot_silhouette()
```

### Analyzing Results

```{r hclust-summary-auto}
results <- hc_auto$summary()

# Quality metrics
cat("Average silhouette:", results$avg_silhouette, "\n")
cat("Variance explained:", results$total_var_explained, "\n")

# Cluster summary
results$clust_summary
```

```{r hclust-members-auto}
# Cluster members
results$clust_members
```

```{r hclust-corr-auto}
# Centroid correlations
results$centroids_correlations
```

### Prediction on Illustrative Variables

```{r hclust-predict-auto}
# Predict cluster for illustrative variables
prediction <- hc_auto$predict(autos_test)

# Labels
prediction$labels

# Proximities to all clusters
prediction$proximities
```

---

## 1.3 Example 2: Qualitative Variables (Canines Dataset)

```{r hclust-canines}
# Load canines dataset
data(canines)

# Display
summary(canines)
head(canines)
```

### Full Pipeline

```{r hclust-fit-canines}
# Select active variables
canines_actives <- canines[, 2:7]

# Initialize for qualitative variables
hc_canines <- HClustVar$new(
  vartype = "qual",
  cah.method = "ward.D2"
)

# Fit
hc_canines$fit(canines_actives)

# Dendrogram
hc_canines$plot_dendrogram()
```

```{r hclust-elbow-canines}
# Aggregation levels
hc_canines$plot_agg_levels()
```

```{r hclust-cut-canines}
# Cut and analyze
hc_canines$cut_tree(4)
canines_result <- hc_canines$summary()

# Results
canines_result$clust_summary
canines_result$clust_members
```

```{r hclust-mds-canines}
# MDS visualization
hc_canines$mds_projection()
```

---

# Part 2: K-means Variable Clustering (KmeansVariables)

## 2.1 Introduction

The `KmeansVariables` class implements K-means clustering of variables using PCA to compute cluster centroids. It groups variables by maximizing their correlation with these centroids.

### Algorithm Overview

1. **Initialization**: Randomly assign variables to K clusters
2. **Centroid Calculation**: Compute first principal component (PCA) as centroid
3. **Distance Calculation**: Compute distance based on correlation with centroids
4. **Reallocation**: Reassign variables to nearest centroid
5. **Iteration**: Repeat until convergence

### Distance Metrics

- **R² (r_squared)**: Groups variables with high absolute correlation
  $$d(x_j, c_k) = \sqrt{1 - r^2(x_j, c_k)}$$

- **Signed r (r_signed)**: Groups only positively correlated variables
  $$d(x_j, c_k) = 1 - r(x_j, c_k)$$

### Key Features

- Multiple random initializations
- Automatic K selection (elbow, silhouette, Calinski-Harabasz)
- Comprehensive diagnostics
- Prediction on illustrative variables

---

## 2.2 Example: FIFA Players Dataset

### Loading and Exploring

```{r kmeans-load}
# Load players dataset
data(players)

# Display structure
cat("Dimensions:", dim(players), "\n")
cat("Variables:", paste(head(names(players), 10), collapse = ", "), "...\n")

# Summary
summary(players[, 1:6])
```

### Fitting the Model

```{r kmeans-fit}
# Initialize KmeansVariables
km <- KmeansVariables$new(
  n_clusters = 5,
  n_init = 10,
  distance_metric = "r_squared",
  random_state = 42
)

# Fit
km$fit(players)

# Print
print(km)
```

### Accessing Results

```{r kmeans-results}
# Cluster assignments
head(km$labels, 10)

# Cluster sizes
km$cluster_sizes

# Total inertia
cat("Total inertia:", round(km$inertia, 4), "\n")
```

---

## 2.3 Visualizations

### Elbow Plot
Elbow plot require a lot of computation time because of all the iterations needed to compute the graph. this method should be avoided on dataset containing a lot of columns.

```{r kmeans-elbow}
km$plot_elbow(k_range = 2:10)
```

### Variable Projection

```{r kmeans-projection}
km$plot_projection(show_labels = TRUE)
```

### Top Contributing Variables

```{r kmeans-contributions}
km$plot_contributions(top_n = 5)
```

---

## 2.4 Analyzing Results

### Summary

```{r kmeans-summary}
results_km <- km$summary(print_summary = TRUE)
```

### Cluster Summary

```{r kmeans-clust-summary}
results_km$clust_summary
```

### Variable Assignments

```{r kmeans-members}
head(results_km$clust_members, 10)
```

### Centroid Correlations

```{r kmeans-centroids}
round(results_km$centroids_correlations, 3)
```

---

## 2.5 Advanced Features

### Automatic K Selection
The automatic selection uses the elbow plot on silhouette to choose the optimal K. Thus, it is requiring a lot of computation time to produce the results.

```{r kmeans-auto}
# Automatic selection using silhouette
km_auto <- KmeansVariables$new(
  n_clusters = "auto",
  k_range = 2:10,
  selection_method = "silhouette",
  random_state = 42
)

km_auto$fit(players)
cat("Optimal K:", km_auto$n_clusters, "\n")
```

### Prediction

```{r kmeans-predict}
# Create test variables
test_vars <- data.frame(
  test_pace = players$pace + rnorm(nrow(players), 0, 3),
  test_defending = players$defending + rnorm(nrow(players), 0, 3)
)

# Predict
km$predict(test_vars)
```

---

# Part 3: Modality Clustering (ModCluster)

## 3.1 Introduction

The `ModCluster` class implements hierarchical clustering of modalities (categories) using the Dice distance coefficient. It clusters the categories of categorical variables rather than the variables themselves.

### Algorithm Overview

1. **Complete Disjunctive Table**: Convert categorical data to binary indicators
2. **Dice Distance**: Compute dissimilarity between all pairs of modalities
3. **Hierarchical Clustering**: Apply agglomerative clustering
4. **MCA Visualization**: Project modalities in reduced dimensional space

### Dice Distance Formula

$$D(j, j') = \frac{1}{2} \sum_{i=1}^{n} (x_{ij} - x_{ij'})^2$$

### Key Features

- Specialized for categorical data
- Multiple linkage methods
- MCA integration for visualization
- Prediction for new observations

---

## 3.2 Example: Congressional Voting Dataset

### Loading and Exploring

```{r modclust-load}
# Load vote dataset
data(vote)

# Display
cat("Dimensions:", dim(vote), "\n")
cat("Variables:", paste(names(vote), collapse = ", "), "\n")

# Summary
summary(vote)
```

### Splitting Data

```{r modclust-split}
# Split for prediction example
set.seed(42)
train_idx <- sample(nrow(vote), size = round(0.8 * nrow(vote)))
vote_train <- vote[train_idx, ]
vote_test <- vote[-train_idx, ]

cat("Training:", nrow(vote_train), "observations\n")
cat("Test:", nrow(vote_test), "observations\n")
```

### Fitting the Model

```{r modclust-fit}
# Initialize ModCluster
mc <- ModCluster$new(
  method = "hclust",
  n_dimensions = 2,
  hclust_method = "average"
)

# Fit
mc$fit(vote_train)

# Print
print(mc)
```

---

## 3.3 Determining Optimal K

### Aggregation Heights

```{r modclust-heights}
mc$plot_heights()
```

### Cutting the Tree

```{r modclust-cut}
mc$cut_tree(3)
print(mc)
```

---

## 3.4 Visualizations

### Dendrogram

```{r modclust-dendro}
mc$plot_dendrogram()
```

### MCA Projection

```{r modclust-mca}
mc$plot_mca()
```

---

## 3.5 Analyzing Results

### Summary

```{r modclust-summary}
summary(mc)
```

---

# Method Comparison

## Overview

| Feature | HClustVar | KmeansVariables | ModCluster |
|---------|-----------|-----------------|----------|
| **Clusters** | Variables | Variables | Modalities |
| **Data types** | Quant / Qual / Mixed | Quantitative only | Categorical only |
| **Algorithm** | Hierarchical (AHC) | Iterative reallocation | Hierarchical (AHC) |
| **Distance** | R, R², Cramer's V | R, R² | Dice coefficient |
| **Visualization** | Dendrogram, MDS | PCA projection | Dendrogram, MCA |
| **Output** | Variable groups | Variable groups | Category groups |

## When to Use Each Method

### HClustVar
- **Best for**: Mixed data, exploring hierarchical structure
- **Pros**: Flexible, handles any variable type, dendrogram visualization
- **Cons**: Must choose where to cut tree

### KmeansVariables
- **Best for**: Large quantitative datasets, when K is known
- **Pros**: Fast, multiple initializations, automatic K selection
- **Cons**: Quantitative only, sensitive to initialization

### ModCluster
- **Best for**: Understanding categorical data patterns
- **Pros**: Specialized for categories, MCA visualization
- **Cons**: Categorical only, many modalities can be hard to interpret

---

# Practical Recommendations

## 1. Data Preparation

- **HClustVar**: Convert characters to factors for qualitative
- **KmeansVariables**: Ensure all numeric, remove constant variables
- **ModCluster**: All variables must be factors

## 2. Choosing K

- Use elbow/aggregation plots
- Check silhouette scores
- Consider interpretability
- Try multiple values

## 3. Quality Assessment

- **Silhouette > 0.5**: Good clustering
- **Silhouette 0.25-0.5**: Reasonable
- **Silhouette < 0.25**: Consider different K or method

## 4. Interpretation

- Examine cluster composition
- Look for thematic patterns
- Use visualizations
- Validate with domain knowledge

---

# Conclusion

This vignette demonstrated the three clustering methods in the VarClustering package:

1. **HClustVar**: Hierarchical variable clustering for quantitative, qualitative, or mixed data
2. **KmeansVariables**: K-means variable clustering with PCA centroids for quantitative data
3. **ModCluster**: Modality clustering with Dice distance for categorical data

Each method provides:
- Cluster quality assessment (silhouette, variance explained)
- Visualization tools
- Prediction capabilities

Choose the appropriate method based on your data type and analysis goals.

---

# References

- Rakotomalala, R. (2022). Clustering de variables. Support de cours, Université Lyon 2. https://eric.univ-lyon2.fr/ricco/cours/slides/classification_de_variables.pdf

- Rakotomalala, R. (2008). Classification de variables. Tutoriels Tanagra pour le Data Mining. http://tutoriels-data-mining.blogspot.com/2008/03/classification-de-variables.html

- Vigneau, E., & Qannari, E. M. (2003). Clustering of variables around latent components. Communications in Statistics-Simulation and Computation, 32(4), 1131-1150. https://doi.org/10.1081/SAC-120023882

- Chavent, M., Kuentz-Simonet, V., Labenne, A., & Saracco, J. (2012). ClustOfVar: An R package for the clustering of variables. Journal of Statistical Software, 50(13), 1-16. https://doi.org/10.18637/jss.v050.i13

- Ward, J. H. (1963). Hierarchical grouping to optimize an objective function. Journal of the American Statistical Association, 58(301), 236-244.

- Lebart, L., Morineau, A., & Piron, M. (2006). Statistique exploratoire multidimensionnelle : Visualisation et inférences en fouille de données (4e éd.). Dunod.

---

# Session Info

```{r session-info}
sessionInfo()
```
