---
title: "Variable Clustering"
author: "Eugénie Barlet, Modou Mboup, Axel Cano"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Variable Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)
```

# Introduction

This vignette demonstrates the use of clustering for variable. We'll explore three main approaches:

1. **Hierarchical Variable Clustering (HClustVar)** - Groups variables based on similarity. This similarity is computed using correlation or squared correlation between quantitative variables, Cramer's V between qualitative variables, and a discretization algorithms to transform quantitatives into qualitatives on a mixed scenarios.
2. **K-Means Clustering (KMeans)** - Focused onto clustering quantitative variables using PCA to compute centroid of clusters.
3. **Quantitative Variable Clustering** - *TODO*

## Loading the library

```{r libraries}
library(VarClustering)  
```

## Dataset Overview

We'll work with three datasets:

- **Auto2005Non4x4.xls**: Mixed variables (quantitative and qualitative) from automotive data
- **players_22_cleaned.csv**: Quantitative variables from FIFA 22 player statistics
- **races_canines.xls**: Qualitative variables on dogs characteristics.

---

# Part 1: Hierarchical Variable Clustering (HClustVar)

## 1.1 Introduction to HClustVar

The `HClustVar` class implements hierarchical clustering of variables using Agglomerative Hierarchical Clustering (AHC). It groups variables based on their similarity, supporting:

- **Quantitative variables**: Using correlation (r) or squared correlation (r²)
- **Qualitative variables**: Using Cramer's V
- **Mixed variables**: Discretization + Cramer's V

### Key Features

- Quantitative / Qualitative and mixed variable types accepted.
- Multiple distance metrics and clustering methods
- Comprehensive diagnostics (silhouette, MDS, dendrograms)
- Cluster quality assessment
- Prediction of illustrative variables

---

## 1.2 Example 1: Quantitative (Automotive Data)

### Loading and Exploring the Data

```{r load-auto-data}
# Load the autos
data(autos)

# Display structure
str(autos)

# Summary statistics
summary(autos)

# Check variable types
sapply(autos, class)
```

### Splitting the data set into train / test

```{r fit-auto-hclust}
autos_train <- autos[, 2:9]
autos_test <- autos[, 10:12]

# Initialize HClustVar
hc_auto <- HClustVar$new(
  vartype = "quant",        # Work on quantitative variables (auto would have been ok as we only have quantitatives).
  dist.metric = "rsquare",
  cah.method = "ward.D2"    # Ward's method (recommended)
)
```

### Fit the model
```{r}
# Fit the model
hc_auto$fit(autos_train)

# Print the object.
print(hc_auto)
```
### Visualizing the Dendrogram

```{r dendrogram-auto}
# Plot dendrogram, possibility to use k to highlight clusters.
hc_auto$plot_dendrogram()
```

### Determining Optimal Number of Clusters

```{r optimal-k-auto}
# Use the elbow method to find optimal k
hc_auto$plot_agg_levels()
```


### Cutting the Tree

```{r cut-tree-auto}
# Cut the tree into 3 clusters (adjust based on elbow plot)
hc_auto$cut_tree(k = 3)
```
### Visualizing the clusters.

```{r}
# Use K to test the clustering on the dendogram.
hc_auto$plot_dendrogram(k = 3)
```
```{r}
# Compute the mds projection to observe the distances between variables.
hc_auto$mds_projection()
```

```{r}
# Display the silhouette plot
hc_auto$plot_silhouette()
```

### Analyzing Clustering Results

```{r summary-auto}
results <- hc_auto$summary()

# Average silhouette of the clustering.
print(paste("Average silhouette:", results$avg_silhouette))
print(paste("Variance explained by the clusters:", results$total_var_explained))
```
```{r}
# Display the cluster summary
results$clust_summary
```
```{r}
# Display the cluster members details.
results$clust_members
```
```{r}
# Display the relationship between centroids
results$centroids_correlations
```
```{r}
# Display the proximity of all variables with all clusters.
results$all_clusters_R2
```
### Prediction on illustrative variables
```{r}
# Make the prediction directly on a dataframe (or matrix) object containing the illustratives.
prediction <- hc_auto$predict(autos_test)

# Extract the labels using the "labels" object
prediction$labels
```
```{r}
# More details on the proximities to the other clusters (r or r_square depending of the selected metric when instanciating the class).
prediction$proximities
```

---

## 1.3 Example 2: Qualitative Variables (canines dataset)

The clustering using HClustVar work also on qualitative variables, using Cramer V as similiraty score.

### Full pipeline
```{r}
# Load the canines dataset.
data(canines)

summary(canines)
head(canines)
```


```{r load-fifa-data}
# Select the active variables.
canines_actives <- canines[, 2:7]

# Initialisation of the clustering class. No "dist.metric" required for qualitative.
hc_canines <- HClustVar$new(
  vartype = "qual",        # Work on qualitative variables.
  cah.method = "ward.D2"    # Ward's method (recommended)
)

# Training the clustering class.
hc_canines$fit(canines_actives)

# Compute the dendrogram.
hc_canines$plot_dendrogram()
```
```{r}
# Compute th eaggregation level screeplot.
hc_canines$plot_agg_levels()
```
```{r}
# observe in the dendrogram
hc_canines$plot_dendrogram(4)
```
```{r}
# cut the tree
hc_canines$cut_tree(4)

# Display the summary
canines_result <- hc_canines$summary()

canines_result$clust_summary
```

```{r}
canines_result$clust_members
```
```{r}
canines_result$centroids_correlations
```
```{r}
canines_result$all_clusters_R2
```
```{r}
# MDS visualisation
hc_canines$mds_projection()
```

This study could be pushed further and used in tandem with modalities clustering (part 3).

---

# Part 2: Kmeans on variable Clustering.

## 1.1 Introduction to KmeansVariables

The `KmeansVariables` class implements kmeans clustering of variables using PAC to generate latent composant as the centroids of clusters. It then groups variables based on their correlation or squared correlation.

This class only implements **quantitative variables** clustering.


### Key Features

- Quantitative clustering.
- R and R² for similarity measure and cluster fusion.
- Comprehensive diagnostics
- Cluster quality assessment
- Prediction of illustrative variables

---

# Part 3: Modalities clustering

## 1.1 Introduction to KmeansVariables



### Key Features





---

# Conclusion

This vignette demonstrated:

1. **HClustVar manipulation**: Initialization, fitting, and interpretation.

---

# Session Info

```{r session-info}
sessionInfo()
```
