---
title: "Complexity_tests"
output: html_document
---

## Introduction

This notebook purpose is to test the complexity of the algorithms developped in this package.
All algorithm will be tested with incremental size datasets (both rows and columns).
All parameters were set up to their default values.
Compexity evaluated on the `$fit()` method only.

## Load data
```{r}
data(autos)
data(vote)
library(ggplot2)
library(dplyr)
```


## Initialization Result dataframe
```{r}
compute_time <- function(obj, data, algorithm, type, growth) {
  
  t0 <- Sys.time()
  fit_result <- obj$fit(data)
  fit_time <- Sys.time() - t0
  fit_time <-  as.numeric(fit_time, units = "secs")

  result <- data.frame(
  "algorithm" = algorithm,
  "type"  = type,
  "n_observations" = nrow(data),
  "n_variables" = ncol(data),
  "fit_time" = fit_time,
  "growth" = growth
  )
  
  return(result)
}

# Initialize global result dataframe.
results <- data.frame(
  "algorithm" = c(),
  "type"  = c(),
  "n observations" = c(),
  "n variables" = c(),
  "fit_time" = c()
)
```

## Quantitative data
```{r}
# Class instanciation.
hcv <- HClustVar$new(vartype = "quant", cah.method="ward.D2")
kmv <- KmeansVariables$new()

df <- NULL


# Complexity in n (observation)
for (i in seq(10)) {
  
  if (i == 1) {
    df <- autos[, 2:9]
  } else {
    df <- rbind(df, df)
  }
  
  result1 <- compute_time(hcv, df, "HClustVar", "quantitative", "rows")
  result2 <- compute_time(kmv, df, "KMeans", "quantitative", "rows")
  
  # Append the result to the global dataframe
  results <- rbind(results, result1, result2)
}

# Complexity in p (variables)
for (i in seq(10)) {
  
  if (i == 1) {
    df <- autos[, 2:9]
  } else {
    df <- cbind(df, df)
  }
  
  
  result1 <- compute_time(hcv, df, "HClustVar", "quantitative", "cols")
  result2 <- compute_time(kmv, df, "KMeans", "quantitative", "cols")
  
  # Append the result to the global dataframe
  results <- rbind(results, result1, result2)
}

```

## Qualitative data

```{r}
# Class instanciation.
hcv <- HClustVar$new(vartype = "qual", cah.method="ward.D2")
mc <- ModCluster$new()


df <- NULL

# Complexity in n (observation)
for (i in seq(10)) {
  
  if (i == 1) {
    df <- vote
  } else {
    df <- rbind(df, df)
  }
  
  result1 <- compute_time(hcv, df, "HClustVar", "qualitative", "rows")
  result2 <- compute_time(mc, df, "ModClust", "qualitative", "rows")
  
  # Append the result to the global dataframe
  results <- rbind(results, result1, result2)
}

# Complexity in p (variables)
for (i in seq(7)) {
  
  if (i == 1) {
    df <- vote
  } else {
    df <- cbind(df, df)
  }
  
  result1 <- compute_time(hcv, df, "HClustVar", "qualitative", "cols")
  result2 <- compute_time(mc, df, "ModClust", "qualitative", "cols")
  
  # Append the result to the global dataframe
  results <- rbind(results, result1, result2)
}
```

```{r}
# Save the data file
write.csv(results, file="complexity.csv")
```

## Plotting graph : Complexity in rows

```{r}
# Read the file if needed
results <- read.csv("complexity.csv")
```


```{r}

plot_time <- function(algorithm, type, growth) {
  
  ggplot(results[results$algorithm == algorithm & results$type == type & results$growth == growth, ], aes(
  x = ifelse(growth == "rows", n_observations, n_variables),
  y = fit_time
)) +
  geom_point() +
  geom_line() +
  labs(
    title = paste("complexity for", algorithm, "on", type, ":", growth, "complexity"),
    x = paste(growth, "size"),
    y = "Fit time (s)"
  )
  
}

for (method in c("HClustVar", "KMeans", "ModClust")) {
  for (type in c("quantitative", "qualitative")) {
    
    if (!(type %in% results$type[results$algorithm == method])) next
    
    for (growth in c("rows", "cols")) {
      print(plot_time(method, type, growth))
    }
  }
}

```
```{r}
grouped_results <- results %>%
  mutate(size = if_else(growth == "rows", n_observations, n_variables)) %>%
  group_by(algorithm, type, growth) %>%
  summarise(
    size = max(size, na.rm = TRUE),
    fit_time = max(fit_time, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(fit_time = round(fit_time, 3))

grouped_results
```
## Conclusion

**Row growth**: Unsurprisingly, increasing the number of rows has minimal impact on performance (linear complexity for all algorithms). Since our algorithms cluster variables (columns), adding more rows has negligible effect.

**Column growth**: HClust-based algorithms (HClustVar and ModClust) exhibit quadratic complexity. ModClust is twice as efficient as HClustVar because the latter also computes Cramér's V (which itself has quadratic complexity). For qualitative variables, processing time increases significantly with the number of columns, as the distance matrix (on which all calculations are based) grows accordingly.

**Comparison with KMeans**: The comparison with KMeans on quantitative variables is insightful. On medium-sized datasets, HClust computes faster than KMeans. However, KMeans shows linear complexity, suggesting a higher computational overhead (due to its stochastic process with multiple initializations and iterations). On larger datasets, HClustVar may become overwhelmed, whereas KMeans should still handle bigger datasets in reasonable time.




